---
title: "Pairwise scatter plots"
author: "Charlotte Soneson, Michael Stadler"
format:
    html:
        toc: true
        toc-location: left
        embed-resources: true
        link-external-icon: true
        code-fold: true
editor_options: 
    chunk_output_type: console
---

## Summary

This document illustrates how to use `GGally` (a companion package to `ggplot2`) to:

- make pairwise scatter plots of columns from a data set
- customize the default panels


## Prepare data

Run the following code to load the packages and data used in this document:

```{r}
#| label: prepare_data
#| code-fold: false

suppressPackageStartupMessages({
    library(tibble)
    library(ggplot2)
    library(GGally)
    library(swissknife)
})

loadExampleData("mycars")

tibble(mycars)
```

## Create basic pairs plot

We start by using the `GGally::ggpairs()` function to create a pairs plot with 
mostly default settings, displaying a subset of the columns from `mycars`.

```{r}
#| label: plot
#| fig-width: 7
#| fig-height: 7

ggpairs(mycars, 
        columns = c("mpg", "disp", "hp", "wt", "transmission")) + 
    theme_bw(13)
```

## Add aesthetics to the panels

Next, we note that we can add mappings (aesthetics) to the panels, e.g. to 
color by a variable.

```{r}
#| label: plot_color
#| fig-width: 7
#| fig-height: 7

ggpairs(mycars, 
        columns = c("mpg", "disp", "hp", "wt"),
        mapping = aes(colour = transmission)) + 
    theme_bw(13)
```

## Customize individual panels

Specifying the mapping as above modifies all the panels in the pairs plot. 
We can also customize individual panels. Here we illustrate this by coloring 
the points in the panels below the diagonal and adding a regression line for 
each group, but displaying only the overall correlation in the panels 
above the diagonal. 

```{r}
#| label: plot_adapted
#| fig-width: 7
#| fig-height: 7

## Define plot function for the scatter plots
plotpoints <- function(data, mapping, ...) {
    ggplot(data = data, mapping = mapping) +
        aes(color = transmission) + 
        geom_smooth(method = "lm", se = FALSE, 
                    formula = "y ~ x",
                    linetype = "dashed") + 
        geom_point(alpha = 0.5, size = 3) +
        theme_bw(13)
}

## Use this function in the ggpairs call
ggpairs(mycars, 
        lower = list(continuous = plotpoints),
        columns = c("mpg", "disp", "hp", "wt"),
        progress = FALSE) + 
    theme_bw(13) + 
    labs(title = "A subset of the mtcars data set",
         subtitle = "Lines represent linear regression")
```

We can also modify the panels above the diagonal to display the overall 
correlation, but change the font size and add a background color according to 
the strength of the correlation.

```{r}
#| label: plot_adapted_correlation
#| fig-width: 7
#| fig-height: 7

## Define correlation function for the panels above the diagonal
cor_fcn <- function(data, mapping, ...) {
    ## Get data
    xData <- GGally::eval_data_col(data, mapping$x)
    yData <- GGally::eval_data_col(data, mapping$y)
    
    ## Calculate correlation
    corr <- cor(xData, yData, method = "pearson")
    
    ## Define background color
    ## The colorRamp() function creates a function that map the interval 
    ## [0, 1] to colors
    if (corr >= 0) {
        cols <- hcl.colors(n = 11, palette = "RdBu")[6:2]
        col <- rgb(colorRamp(cols)(abs(corr)),
                   maxColorValue = 255)
    } else {
        cols <- hcl.colors(n = 11, palette = "RdBu")[6:10]
        col <- rgb(colorRamp(cols)(abs(corr)),
                   maxColorValue = 255)
    }

    ## Construct plot
    ggplot(data = data, mapping = mapping) +
        annotate(x = 0.5, y = 0.5, 
                 label = paste0("Corr: ", round(corr, digits = 3)),
                 geom = "text",
                 size = abs(corr) * 5 + 1) +
        theme_bw(13) + xlim(c(0, 1)) + ylim(c(0, 1)) +
        theme(panel.background = element_rect(fill = col),
              panel.grid.major = element_blank(),
              panel.grid.minor = element_blank())
}

ggpairs(mycars, 
        lower = list(continuous = plotpoints),
        upper = list(continuous = cor_fcn),
        columns = c("mpg", "disp", "hp", "wt"),
        progress = FALSE) + 
    labs(title = "A subset of the mtcars data set",
         subtitle = "Lines represent linear regression")
```

## Remarks

* It is possible to define customized plots separately for continuous, 
discrete and 'combo' data types, see the [ggpairs documentation](https://ggobi.github.io/ggally/articles/ggpairs.html) for more 
details. 

## Session info

```{r}
#| label: session_info
sessioninfo::session_info()
```
