---
title: "Combining plots in a grid"
author: "Charlotte Soneson, Michael Stadler"
format:
    html:
        toc: true
        toc-location: left
        embed-resources: true
        link-external-icon: true
        code-fold: true
editor_options: 
    chunk_output_type: console
---

## Summary

This document illustrates how to use `cowplot` (a companion package to `ggplot2`) to:

- combine multiple panels into a single figure
- include a non-ggplot2 panel and set a joint title for the figure


## Prepare data

Run the following code to prepare the data used in this document:

```{r}
#| label: prepare_data
#| code-fold: false

suppressPackageStartupMessages({
    library(tibble)
})

# built-in `mtcars` dataset (see ?mtcars)
# ... convert some columns to factors
mycars <- mtcars
mycars$model <- rownames(mycars)
mycars$cyl <- factor(mycars$cyl, levels = c("4","6","8"))
mycars$engine_shape <- factor(c("0" = "V-shaped", "1" = "straight")[as.character(mycars$vs)])
mycars$transmission <- factor(c("0" = "automatic", "1" = "manual")[as.character(mycars$am)])

tibble(mycars)
```


## Create figure

### Load packages
```{r}
#| label: load_packages

library(ggplot2)
library(cowplot)
```


### Plot

We first create the individual plots that we will later use as panels in the 
combined plot. 

```{r}
#| label: plot
#| fig-width: 5
#| fig-height: 5

p1 <- ggplot(data = mycars, mapping = aes(x = mpg, y = hp)) +
    geom_point(size = 3, alpha = 0.5, aes(color = transmission)) +
    labs(x = "Fuel efficiency (miles/gallon)", y = "Gross horsepower") +
    theme_bw(13) +
    theme(panel.grid = element_blank(),
          legend.position = "bottom")
p1
p2 <- ggplot(data = mycars, mapping = aes(x = mpg)) +
    geom_histogram(bins = 10, 
                   color = "darkgrey", fill = "lightgrey") +
    labs(x = "Fuel efficiency (miles/gallon)") +
    theme_bw(13) +
    theme(panel.grid = element_blank(),
          legend.position = "bottom")
p2
p3 <- ggplot(data = mycars, mapping = aes(x = transmission, y = hp)) + 
    geom_boxplot(outlier.size = -1) + 
    geom_jitter(width = 0.2, height = 0, alpha = 0.5) + 
    labs(x = "Transmission", y = "Gross horsepower") + 
    theme_bw(13) + 
    theme(panel.grid = element_blank(),
          legend.position = "bottom")
p3
```

Next we combine these plots into a single figure, including also a jpg image 
of a puppy (which does not get a panel label). We set a single title for the 
full plot. 

```{r}
#| label: plot_combine
#| fig-width: 7
#| fig-height: 7

## Generate image object from jpg file
img <- ggdraw() + 
    draw_image("puppy-dog-1394586170ZjL.jpg")

## Create a joint title
jointtitle <- ggdraw() + 
    draw_label("Combined plot title",
               fontface = "bold", x = 0, hjust = 0)

plot_grid(jointtitle, 
          plot_grid(
              p3, p1, img, p2, 
              ncol = 2, labels = c("A", "B", "", "C"), 
              align = "hv", axis = "tblr", rel_widths = c(0.5, 1)
          ), 
          ncol = 1, rel_heights = c(0.1, 1)
)
```


## Remarks

* If panels have a shared legend and we only wish to include it once, we 
can extract it using `cowplot::get_legend()` and include it as a separate panel.
* To include `ggpairs` panels, see the tips [here](https://fmicompbio.github.io/intermediate_ggplot2_FMI2023/intermediate_ggplot2.html#combining-plot-panels).

## Session info

```{r}
#| label: session_info
sessioninfo::session_info()
```
