---
title: "Placing legends"
author: "Michael Stadler, Charlotte Soneson"
format:
    html:
        toc: true
        toc-location: left
        embed-resources: true
        link-external-icon: true
        code-fold: true
editor_options: 
    chunk_output_type: console
---

## Summary

This document illustrates how to place legends in different positions in a 
`ggplot`.


## Prepare data

Run the following code to load the data used in this document:

```{r}
#| label: prepare_data
#| code-fold: false

suppressPackageStartupMessages({
    library(dplyr)
    library(tibble)
    library(swissknife)
    library(ggplot2)
})

loadExampleData("mycars")

tibble(mycars)
```


## Step 1 - construct basic plot

```{r}
#| label: make_basic_plot

gg <- ggplot(mycars, 
             aes(x = mpg, y = disp, color = cyl)) + 
    geom_point(size = 4) +
    scale_color_manual(values = c(`4` = "purple", `6` = "orange", 
                                  `8` = "forestgreen"), 
                       name = "Number of\ncylinders") + 
    labs(x = "Miles/gallon", y = "Displacement") + 
    theme_bw(base_size = 13)
gg
```

## Step 2 - move legend to the bottom

```{r}
#| label: legend_bottom

gg + theme(legend.position = "bottom")
```

## Step 3 - move legend inside the canvas (top right corner)

```{r}
#| label: legend_inside

gg + theme(legend.position = "inside",
           legend.position.inside = c(0.85, 0.85),
           legend.background = element_rect(color = "black"),
           legend.title.position = "left")
```

## Step 4 - hide specific legend

Use the `cyl` column for both color and shape, but only show the legend for 
the shape. 

```{r}
#| label: hide_one_legend

ggplot(mycars, 
       aes(x = mpg, y = disp, color = cyl, shape = cyl)) + 
    geom_point(size = 4) +
    scale_color_manual(values = c(`4` = "purple", `6` = "orange", 
                                  `8` = "forestgreen"), 
                       name = "Number of\ncylinders") + 
    labs(x = "Miles/gallon", y = "Displacement") + 
    theme_bw(base_size = 13) + 
    guides(color = "none", 
           shape = guide_legend(title = "Number of\ncylinders"))
```

## Step 5 - place different legends in different positions

```{r}
ggplot(mycars, 
       aes(x = mpg, y = disp, color = gear, shape = cyl)) + 
    geom_point(size = 4) +
    scale_color_continuous(name = "Number of\ngears") + 
    scale_shape_discrete(name = "Number of\ncylinders") + 
    labs(x = "Miles/gallon", y = "Displacement") + 
    theme_bw(base_size = 13) + 
    guides(color = guide_colorbar(position = "right"), 
           shape = guide_legend(position = "bottom"))
```

## Links

* Legends in `ggplot2` 3.5.0: [https://www.tidyverse.org/blog/2024/02/ggplot2-3-5-0-legends/](https://www.tidyverse.org/blog/2024/02/ggplot2-3-5-0-legends/)

## Session info

```{r}
#| label: session_info
sessioninfo::session_info()
```
