---
title: "Intermediate ggplot2 - tips and tricks"
author: "Charlotte Soneson, Michael Stadler"
date: "`r Sys.Date()`"
format: 
    html:
        toc: true
editor_options: 
  chunk_output_type: console
---


# Load packages

```{r}
suppressPackageStartupMessages({
    library(ggplot2)
    library(ggforce)
    library(ggalt)
    library(ggridges)
    library(GGally)
    library(patchwork)
    library(cowplot)
    library(scales)
    library(forcats)
    library(plotly)
    library(ggiraph)
    library(ggsignif)
    library(ggstatsplot)
    library(gghighlight)
    library(ggrepel)
    library(ggtext)
    library(ggpackets)
    library(RColorBrewer)
    library(circlize)
    library(sessioninfo)
    library(dplyr)
    library(fueleconomy)
    library(tidyr)
    library(ggnewscale)
})
```

# Quick look at the data

```{r}
(df <- vehicles %>%
     dplyr::filter(year >= 2000))
```

# Start with a basic plot

```{r}
ggplot(df, aes(x = fuel)) + 
    geom_bar() + 
    coord_flip()
```

# Changing order of factor levels

Use the `forcats` package.

## Order by frequency - `fct_infreq`

```{r}
ggplot(df, aes(x = fct_infreq(fuel))) + 
    geom_bar() + 
    coord_flip()
```

## Order by the order they first appear in the data - `fct_inorder`

```{r}
ggplot(df, aes(x = fct_inorder(fuel))) + 
    geom_bar() + 
    coord_flip()
```

## Lump together rare factors - `fct_lump_n`

```{r}
ggplot(df, aes(x = fct_infreq(fct_lump_n(fuel, n = 3, other_level = "Other")))) + 
    geom_bar() + 
    coord_flip()
```

## Reorder one factor by the values of another - `fct_reorder`

I use this e.g. to plot library sizes of samples, ordered by experiment. 

```{r}
ggplot(df %>% group_by(class) %>% summarize(mean_hwy = mean(hwy)), 
       aes(x = fct_reorder(class, mean_hwy), y = mean_hwy)) + 
    geom_col() + 
    coord_flip()

ggplot(df %>% group_by(class) %>% summarize(mean_hwy = mean(hwy)), 
       aes(x = fct_reorder(class, mean_hwy), y = mean_hwy)) + 
    geom_point() + 
    coord_flip()
```

# Faceting

## `facet_wrap` 

```{r}
ggplot(df, aes(x = hwy, y = cty)) +
    geom_point() + 
    facet_wrap(~ fct_lump_n(fuel, n = 3, other_level = "Other"))
```

```{r}
ggplot(df, aes(x = hwy, y = cty)) +
    geom_point() + 
    facet_wrap(~ fct_lump_n(fuel, n = 3, other_level = "Other"), 
               scales = "free")
```

## `facet_grid`

```{r}
ggplot(df, aes(x = hwy, y = cty)) +
    geom_point() + 
    facet_grid(fct_lump_n(class, n = 3) ~ fct_lump_n(fuel, n = 3, other_level = "Other"))
```

## Showing all the data as 'background' in each facet

```{r}
ggplot(df, aes(x = hwy, y = cty)) +
    geom_point(data = transform(df, fuel = NULL), colour = "grey80") +
    geom_point() + 
    facet_wrap(~ fct_lump_n(fuel, n = 3, other_level = "Other"), 
               scales = "free")
```

```{r}
## gghighlight
ggplot(df %>% mutate(fuel_lumped = fct_lump_n(fuel, n = 3, other_level = "Other")),
       aes(x = hwy, y = cty)) +
    geom_point(aes(color = fuel_lumped)) + 
    gghighlight() + 
    facet_wrap(~ fuel_lumped, scales = "free")

ggplot(df %>% mutate(fuel_lumped = fct_lump_n(fuel, n = 3, other_level = "Other")),
       aes(x = hwy, y = cty)) +
    geom_point(aes(color = fuel_lumped)) + 
    gghighlight(fuel_lumped == "Premium")

ggplot(df %>% mutate(fuel_lumped = fct_lump_n(fuel, n = 3, other_level = "Other")),
       aes(x = hwy, y = cty)) +
    geom_point() + 
    gghighlight(hwy - cty > 15 | ((abs(hwy - 60) < 10) & (abs(cty - 75) < 10)), 
                label_key = make)
```



## Facetting bar plots with different numbers of categories in each facet

```{r, fig.width = 10}
## facet_wrap with scales = "fixed" -> leave a missing bar for empty categories
ggplot(df, aes(x = fct_lump_n(make, n = 3))) +
    geom_bar() + 
    facet_wrap(~ fct_lump_n(fuel, n = 3, other_level = "Other"), 
               scales = "fixed", nrow = 1)

## facet_wrap with scales = "free" -> don't leave empty space, 
## but bars will be of different width
ggplot(df, aes(x = fct_lump_n(make, n = 3))) +
    geom_bar() + 
    facet_wrap(~ fct_lump_n(fuel, n = 3, other_level = "Other"), 
               scales = "free", nrow = 1)

## facet_grid lets you adjust the space allocated to each facet, but will keep 
## the y-axis the same for all panels
ggplot(df, aes(x = fct_lump_n(make, n = 3))) +
    geom_bar() + 
    facet_grid(~ fct_lump_n(fuel, n = 3, other_level = "Other"), 
               scales = "free", space = "free")

## ggforce::facet_row lets you adjust the space allocated to each facet and use a 
## free y-axis
ggplot(df, aes(x = fct_lump_n(make, n = 3))) +
    geom_bar() + 
    facet_row(~ fct_lump_n(fuel, n = 3, other_level = "Other"), 
               scales = "free", space = "free")
```


# More on bar plots (without faceting)

```{r}
## Default - stacked bar plot
ggplot(df, aes(x = fct_lump_n(fuel, n = 3, other_level = "Other"), 
               fill = fct_lump_n(make, n = 3))) +
    geom_bar()

## Side-by-side bar plot (but bars have different widths)
ggplot(df, aes(x = fct_lump_n(fuel, n = 3, other_level = "Other"), 
               fill = fct_lump_n(make, n = 3))) +
    geom_bar(position = position_dodge())

## Side-by-side bar plot with consistent width (but the same category 
## will not always be in the same relative position)
ggplot(df, aes(x = fct_lump_n(fuel, n = 3, other_level = "Other"), 
               fill = fct_lump_n(make, n = 3))) +
    geom_bar(position = position_dodge2(preserve = "single"))

## Side-by-side bar plot with consistent width and position
ggplot(df %>% mutate(fuel = fct_lump_n(fuel, n = 3, other_level = "Other"), 
                     make = fct_lump_n(make, n = 3)) %>%
           count(fuel, make) %>%
           complete(fuel = unique(fuel), make = unique(make), 
                    fill = list(n = 0)), 
       aes(x = fuel, y = n, fill = make)) + 
    geom_bar(stat = "identity", position = position_dodge2(preserve = "single"))
```


# Changing scales for axes

## Axes with percent (%) values

```{r}
## scales::percent will multiply by 100
ggplot(df, aes(x = fct_lump_n(fuel, n = 3, other_level = "Other"), 
               fill = fct_lump_n(make, n = 3))) +
    geom_bar(position = "fill") + 
    scale_y_continuous(labels = scales::percent) + 
    labs(y = "Percentage")

## scales::percent_format has more arguments
ggplot(df, aes(x = fct_lump_n(fuel, n = 3, other_level = "Other"), 
               fill = fct_lump_n(make, n = 3))) +
    geom_bar(position = "fill") + 
    scale_y_continuous(labels = scales::percent_format(scale = 100, accuracy = 0.01)) + 
    labs(y = "Percentage")
```

## Limit axis range

```{r}
## xlim() + ylim() will replace out-of-range data by NA
ggplot(df, aes(x = hwy, y = cty)) +
    geom_point(size = 8, shape = 1) +
    xlim(11, 55)

## coord_cartesian() just sets the displayed range
ggplot(df, aes(x = hwy, y = cty)) +
    geom_point(size = 8, shape = 1) +
    coord_cartesian(xlim = c(11, 59))
```

# Interactive plots

Be careful - these plots may get very big!

```{r}
## Turn a ggplot into an interactive plot with ggplotly
ggplotly(
    ggplot(df %>% dplyr::filter(year == 2005), aes(x = hwy, y = cty, 
                                                   label = make)) +
        geom_point(),
    tooltip = c("label", "x", "y", "text")
)
```

- `ggiraph`

# Titles, subtitles and captions

```{r}
ggplot(df %>% dplyr::filter(year == 2005), aes(x = hwy, y = cty)) +
    geom_point() + 
    labs(title = "This is the title", 
         subtitle = "And a subtitle",
         caption = "And the caption goes here",
         x = "Highway fuel economy (mpg)",
         y = "City fuel economy (mpg)")
```

Note that not all of these are compatible with interactive plots:

```{r}
ggplotly(
    ggplot(df %>% dplyr::filter(year == 2005), aes(x = hwy, y = cty)) +
        geom_point() + 
        labs(title = "This is the title", 
             subtitle = "And a subtitle",
             caption = "And the caption goes here",
             x = "Highway fuel economy (mpg)",
             y = "City fuel economy (mpg)")
)
```

# Adding statistical information to plots

- `ggsignif`, `ggstatsplot`

# Adding text and annotation

```{r}
## ggforce
ggplot(df %>% dplyr::filter(year == 2005), aes(x = hwy, y = cty)) +
    geom_point() + 
    geom_mark_ellipse(aes(label = "Interesting points", 
                          description = "These points are interesting",
                          filter = (cty - hwy) > 1), 
                      fill = "pink")

## ggalt
ggplot(iris, aes(x = Petal.Width, y = Sepal.Width, colour = Species)) +
    geom_point() + 
    geom_encircle(aes(group = Species))
```

```{r}
## gghighlight
ggplot(diamonds %>% group_by(carat, cut) %>% summarize(mean_price = mean(price)),
       aes(x = carat, y = mean_price, color = cut)) +
    geom_line() + 
    gghighlight(cut %in% c("Good", "Ideal")) + 
    facet_wrap(~ cut)
```


- `geomtextpath`
- `ggtext`
- `ggrepel`
- add number of values in bar plot (https://r-graphics.org/recipe-bar-graph-labels)

# Customizing legends

```{r}
## legend position (outside of plot)
## allowed values for the arguments legend.position are:
## "left", "top", "right", "bottom"
ggplot(df, aes(hwy, cty, color = fuel)) +
    geom_point() +
    theme(legend.position = "bottom")

## number of rows (discrete scales)
ggplot(df, aes(hwy, cty, color = fuel)) +
    geom_point() +
    theme(legend.position = "bottom") +
    guides(color = guide_legend(nrow = 5))

## override graphical parameters
ggplot(diamonds, aes(carat, price, color = cut)) +
    geom_point(size = 1, alpha = 0.1) +
    theme(legend.position = "bottom") +
    guides(color = guide_legend(override.aes = list(size = 4, alpha = 1)))

## move legend inside plot
## c(0,0) corresponds to the “bottom left” and
## c(1,1) corresponds to the “top right” position
ggplot(diamonds, aes(carat, price, color = cut)) +
    geom_point(size = 1, alpha = 0.1) +
    theme(legend.position = c(1, 0),
          legend.justification = c(1, 0)) +
    guides(color = guide_legend(override.aes = list(size = 4, alpha = 1)))
```

# Specifying color palettes

```{r}
## One-sided gradient
ggplot(df %>% dplyr::filter(year == 2005), aes(x = hwy, y = cty, color = cyl)) +
    geom_point() + 
    scale_color_gradient(low = "white", high = "darkblue")

## Two-sided gradient
ggplot(df %>% dplyr::filter(year == 2005), aes(x = hwy, y = cty, color = scale(cyl))) +
    geom_point() + 
    scale_color_gradient2(low = "red", mid = "white", high = "darkblue", 
                          midpoint = 0)

## Custom gradient
ggplot(df %>% dplyr::filter(year == 2005), aes(x = hwy, y = cty, color = scale(cyl))) +
    geom_point() + 
    scale_color_gradientn(colours = hcl.colors(9, "Spectral"))

## viridis continuous colors
## viridis_c - continuous
## viridis_b - binned
## viridis_d - discrete
ggplot(df %>% dplyr::filter(year == 2005), aes(x = hwy, y = cty, color = scale(cyl))) +
    geom_point() + 
    scale_color_viridis_c()

## Manual colors
ggplot(df, aes(x = fct_lump_n(fuel, n = 3, other_level = "Other"), 
               fill = fct_lump_n(make, n = 3))) +
    geom_bar(position = "fill") + 
    scale_fill_manual(values = c(BMW = "darkblue", Chevrolet = "red", 
                                 Ford = "forestgreen", Other = "grey85"),
                      name = "Make")

## Colors from existing palettes
## RColorBrewer palettes
display.brewer.all()
ggplot(df, aes(x = fct_lump_n(fuel, n = 3, other_level = "Other"), 
               fill = fct_lump_n(make, n = 3))) +
    geom_bar(position = "fill") +
    scale_fill_brewer(palette = "Set2", name = "Make")

ggplot(df, aes(x = fct_lump_n(fuel, n = 3, other_level = "Other"), 
               fill = fct_lump_n(make, n = 3))) +
    geom_bar(position = "fill") +
    scale_fill_brewer(palette = "Greens", name = "Make")

## 'viridis' colors
ggplot(df, aes(x = fct_lump_n(fuel, n = 3, other_level = "Other"), 
               fill = fct_lump_n(make, n = 3))) +
    geom_bar(position = "fill") +
    scale_fill_viridis_d()

## multiple scales for color or fill: ggnewscale::new_scale
ggplot(df %>% dplyr::filter(year == 2005), aes(x = hwy, y = cty, color = cyl)) +
    geom_point(size = 3) + 
    scale_color_viridis_c() +
    new_scale("color") +
    geom_point(aes(color = fuel), shape = 21, size = 4, stroke = 1) +
    scale_color_brewer(palette = "Dark2")

## cap color range
## - one way of doing it: limit color range + set color for out-of-range NA values
cols <- hcl.colors(8)
ggplot(diamonds, aes(x = carat, y = depth, color = price)) +
    geom_point() +
    scale_color_gradientn(colors = cols,
                          limits = c(min(diamonds$price), 10000),
                          na.value = cols[length(cols)])

```


# Specifying your own theme

- `ggpackets`

# Combining plot panels

```{r}
g1 <- ggplot(df, aes(x = fct_lump_n(fuel, n = 3, other_level = "Other"), 
                     fill = fct_lump_n(make, n = 3))) +
    geom_bar(position = "fill") +
    scale_fill_brewer(palette = "Set2", name = "Make")
g2 <- ggplot(df %>% dplyr::filter(year == 2005), aes(x = hwy, y = cty, color = scale(cyl))) +
    geom_point() + 
    scale_color_gradient2(low = "red", mid = "white", high = "darkblue", 
                          midpoint = 0)
g3 <- ggplot(df %>% group_by(class) %>% summarize(mean_hwy = mean(hwy)), 
             aes(x = fct_reorder(class, mean_hwy), y = mean_hwy)) + 
    geom_point() + 
    coord_flip()
(g12 <- cowplot::plot_grid(g1, g2 + theme(legend.position = "none"),
                           ncol = 1, labels = c("A", "B"), align = "v"))
(g12 <- cowplot::plot_grid(g1, g2 + theme(legend.position = "none"),
                           ncol = 1, labels = c("A", "B"), align = "v", axis = "lr"))
cowplot::plot_grid(g12, g3, nrow = 1, labels = c("", "C"))
```

```{r}
get_legend(g1)
cowplot::plot_grid(g1 + theme(legend.position = "none"), 
                   g1 + theme(legend.position = "none"), 
                   get_legend(g1), 
                   nrow = 1, rel_widths = c(1, 1, 0.4))
```

```{r}
## patchwork
g1 | g2
g1 / g2 + theme(legend.position = "none")
(g1 / g2) | g3
```


- shared title


# Other material, sources for some of the material presented above

- [https://uc-r.github.io/ggplot](https://uc-r.github.io/ggplot)
- [https://exts.ggplot2.tidyverse.org/gallery/](https://exts.ggplot2.tidyverse.org/gallery/)
- [https://www.youtube.com/watch?v=8ikFe82Mb1I&ab_channel=R-LadiesTunis](https://www.youtube.com/watch?v=8ikFe82Mb1I&ab_channel=R-LadiesTunis)
- [https://stulp.gmw.rug.nl/ggplotworkshop/advancedggplot.html#showing-all-data-in-facets](https://stulp.gmw.rug.nl/ggplotworkshop/advancedggplot.html#showing-all-data-in-facets)
- [https://stackoverflow.com/questions/61922380/preserve-location-of-missing-columns-in-combined-bar-plot](https://stackoverflow.com/questions/61922380/preserve-location-of-missing-columns-in-combined-bar-plot)

# Session info

```{r}
sessioninfo::session_info()
```

